---
title: "Class 08: Mini Project"
author: "Alexandra Garcia (A16278166)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for us to explore a complete analysis using the unsupervised learning techniques covered in class. I'll extend what I’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. 

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data import

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

Make sure we do not include sample id or the diagnosis column in further analysis because you don't want personal patient information to be included in analysis. 

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)

wisc.data <- wisc.df[, -1]

dim(wisc.data)
```

```{r}
head(wisc.data)
```

> Q1. How many observations are in this dataset?

```{r}
dim(wisc.data)
```

There are 569 patients.  

> Q2. How many of the observations have a malignant diagnosis?

```{r}
# you can use table function
table(diagnosis)

# or you can use sum
sum(diagnosis == "M")
```

There are 212 malignant diagnosis. 

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

There are 10 columns titled with _mean. 

## Principal Component Analysis

The main function in base R for PCA is called `prcomp()`. A optiional argument `scale` should nearly always be switched to `scale=TRUE` for this function. 

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
```{r}
# Perform PCA on wisc.data. You want to always scale the data to make sure that every column in your analysis is contributing equally. 
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```

```{r}
# Look at summary of results
summary(wisc.pr)
```
> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% 

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC7

Let's make our main results figure - the "PC plot" or "score plot" or "ordienation" etc. 

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot shows that there is a clear separation between B and M diagnoses, meaning there is something going on somewhere in the data that is determining the diagnosis. 

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```
The values are set lower on the y axis on PC3 but there is still a clear separation between B and M diagnosis. 

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)
```

```{r}
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

The concave.points_mean feature of the tumor features is highly contributing to the malignancy of a tumor because if we look at the PCA plot above of PC1 (x axis), the value -0.26 is going left of the 0 point on the axis, where all the malignant diagnoses are placed. 

## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

data.dist <- dist(data.scaled)

wisc.hclust <- hclust(data.dist, method="complete")
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=17, col="red", lty=2)
```
Height 17 has 4 clusters. 

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, 4)

table(wisc.hclust.clusters, diagnosis)
```
> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust.sing <- hclust(data.dist, method="sing")
plot(wisc.hclust.sing)

wisc.hclust.avg <- hclust(data.dist, method="average")
plot(wisc.hclust.avg)

wisc.hclust.ward <- hclust(data.dist, method="ward.D2")
plot(wisc.hclust.ward)
```
My favorite method is ward.D2 because it allows for a nicer display of the clustering that is not as crowded at the top. 

```{r}
wisc.pr.hclust <- hclust(data.dist, method="ward.D2")

grps <- cutree(wisc.pr.hclust, k=2)
table(grps)

table(grps, diagnosis)

plot(wisc.pr$x[,1:2], col=grps)

plot(wisc.pr$x[,1:2], col=diagnosis)

g <- as.factor(grps)
levels(g)

g <- relevel(g,2)
levels(g)

# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```
```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

> Q13. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
I think that the new model with 4 clusters separates out the data even more, which is better. 

> Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```

This model is better for separating out the diagnosis because it can allows you to check out the true and false positives in each diagnosis group. 