---
title: "Class 10: Halloween Mini-Project"
author: "Alexandra Garcia PID: A16278166"
format: pdf
toc: TRUE
---

## Importing candy data 

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 types of candy. 

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types. 

## What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way Midnight",]$winpercent
```
60%
> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
77% 
> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
50% 

```{r, message=FALSE}
library(dplyr)

candy |> nrow()
```
```{r}
win <- candy$winpercent
win.mean <- mean(win)
round(win.mean)
```

```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent row is on a different scale. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think the 0 and 1 represent false and true. 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(bins=20)
```
> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) + 
  aes(winpercent) +
  geom_density()
```
No, it is not symmetrical. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
The center is above 50 if you use the mean. 

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in dataset (row)
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
choc.candy
# 2. Extract their winpercent values
choc.win <- choc.candy$winpercent
choc.win
# 3. Find the mean of these values
choc.mean <- mean(choc.win)
choc.mean
# 4-6. Do the same for fruit candy
fruit.inds <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.inds,]
fruit.candy

fruit.win <- fruit.candy$winpercent
fruit.win

fruit.mean <- mean(fruit.win)
fruit.mean
# 7. Which mean is higher
choc.mean > fruit.mean
```
Chocolate is more popular than fruity candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
No, it is not statistically significant based on the p value. 

## Overall candy ranking 

> Q13. What are the five least liked candy types in this set?

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ord.ind,],5)
```
Snickers, Kit Kat, Twix, Reese's Minis, Reese's PB cup

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
Add some color

```{r}

my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) +
  aes(x= winpercent, 
      y= reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols) 
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Winpercent vs. Pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12)], n=5 )
```
Reese's Minis

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


## Correlation

```{r}
cij <- cor(candy)
```
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate 

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

## Principal component analysis

The main function in base R is `prcomp()` and we want to set `scale=TRUE`

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Let's look at our first main result figure - the "PC plot" or PC1 vs. PC2

```{r}
pca$x
```
```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
```
Don't forget about your variable "loadings" - how the original variables contribute to your new PCs... 

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Pluribus, hard, fruity are all in the positive direction, which makes sense because they are all characteristics that those candies have (laffy taffy, airheads, warheads). 

